## Problem B: The gamma distribution


### 1. Rejection sampling

## (a) Acceptance probability

The acceptance probability is the inverse of the constant $c$ used in the rejection-sampling algorithm. The constant $c$ is chosen to be the smallest value that satisfies $c \geq \frac{f(x)}{g(x)}$. 

With $f(x) = \frac{1}{\Gamma(\alpha)}x^{\alpha -1}e^{-x}$ and $g(x)$ as specified in A.2, we must choose $c$ such that

$$
c \geq \frac{f(x)}{g(x)} = \begin{cases}\frac{1}{\Gamma(\alpha)}\frac{\alpha + e}{e\alpha}\frac{1}{e^x},\quad 0 < x < 1 \\ 
\frac{1}{\Gamma(\alpha)}\frac{\alpha + e}{e\alpha}x^{\alpha - 1},\quad x \geq 1\end{cases}
$$
This functions attains it's maximum at $x=1$, where we find $\frac{f(x)}{g(x)}\mid_{x=1} = \frac{1}{\Gamma(\alpha)}\frac{\alpha + e}{e\alpha}$ and by choosing $c = \frac{f(x)}{g(x)}\mid_{x=1}$ we satisfy the criterion above. This leads to the acceptance probability $\Gamma(\alpha)\frac{e\alpha}{e + \alpha}$.

## (b) Rejection-sampling implementation


```{r}
n <- 100000
alpha <- 0.4

rejection_sampling_f <- function(n, alpha){
  dg_fixed <- partial(dg, alpha=alpha)
  df <- as_mapper(~1/gamma(alpha)*.x^(alpha - 1)*exp(-.x))
  c <- (alpha + exp(1))/(alpha*exp(1))/gamma(alpha)
  
  n_remaining <- n
  samples <- vector()
  
  while (n_remaining > 0){
    acceptance_threshold_mapper <- as_mapper(~1/c*df(.x)/dg_fixed(.x))
    
    xs <- rg(n_remaining, alpha=alpha) %>% use_series(value)
    acceptance_threshold <- xs %>% map_dbl(acceptance_threshold_mapper)
    
    accepted_xs <- xs %>%
      keep(runif(n_remaining) < acceptance_threshold) %>%
      enframe()
    samples <- c(samples, accepted_xs$value)
    
    n_remaining <- n_remaining - nrow(accepted_xs)
  }
  return(samples)
}

rv_f <- rejection_sampling_f(n, alpha)

df <- as_mapper(~1/gamma(alpha)*.x^(alpha - 1)*exp(-.x))
dg_fixed <- partial(dg, alpha=alpha)
c <- (alpha + exp(1))/(alpha*exp(1))/gamma(alpha)
dg_scaled <- as_mapper(~dg_fixed(.x)*c)

ggplot(rv_f %>% enframe()) +
  geom_histogram(aes(x=value, y=..density..), binwidth=0.05, boundary=0) +
  stat_function(fun=df, color='red', size=1) +
  stat_function(fun=dg_scaled, color='green', size=1, linetype="dashed") +
  ylim(0, 1) +
  xlim(0, NA)

print(paste("Acceptance probability:", 1/c))

```



### 2. Ratio of uniforms method

## (a)

To find the values of $a$ and $b_{+}$ we take the of derivative of $f^{\star}(x)$ and $x^2f^{\star}(x)$and set it equal to zero. This gives us the values for which the two functions are maximized. By evaluating the functions at these points, $x = \sqrt{\alpha - 1}$ and $x = \sqrt{\alpha + 1}$ respectively, one obtains that
$a = \sqrt{\left(\frac{\alpha - 1}{e}\right)^{\alpha - 1}}$ and $b_{+} = \sqrt{\left(\frac{\alpha + 1}{e}\right)^{\alpha + 1}}$. The function $f^{\star}(x)$ is zero for $x \leq 0$, so it's supremum is 0 in this region, giving us $b_{-} = 0$.

Neither the values of $a$ and $b_{+}$ nor the function $f^{\star}(x)$ is easy to evaluate numerically for high values of $alpha$. This problem is solved by sampling directly from the distribution of $Y_1 = \log(X_1)$ and $Y_2 = \log(X_2)$, and proceeding by checking whether the values of $Y_1$ and $Y_2$ is such that $(X_1, X_2}$ falls within the region $C_f$.

The condition on a realization of $Y$ for checking whether $X$ is in $C_f$ becomes 

$$
y_1 = \log(x_1) \leq \frac{1}{2}\left[(\alpha - 1)(\log(x_2) - \log(x_1)) - e^{\log(x_2) - \log(x_1)}\right]
$$
To be able to sample from the distribution of $Y$ we need to find the inverse cumulative distribution function. The distribution function of $Y_1$ is given by $f_{Y_1}(x) = e^{x - \log(a)}\mathrm{1}_{\{x \leq \log(a)\}}$. The cumulative distribution is given by $F_{Y_1}(x) = e^{x - \log(a)}$. Setting $u = F_{Y_1}(x)$ yields $x = \log(u) + \log(a)$. In an equivalent fashion, we obtain $x = \log(u) + \log(b_{+})$ for sampling $Y_2$.

```{r}
alpha <- 500
n <- 10000000

ratio_of_uniforms_f <- function(n, alpha){
  n_remaining <- n
  attempts <- 0
  samples <- vector()
  
  log_a <- (alpha - 1)/2*log((alpha - 1)/exp(1))
  log_b_plus <- (alpha + 1)/2*log((alpha + 1)/exp(1))
  
  while (n_remaining > 0) {
    log_x1 <- runif(n_remaining) %>% log() %>% add(log_a)
    log_x2 <- runif(n_remaining) %>% log() %>% add(log_b_plus)
    random_samples <- tibble(log_x1, log_x2)
  
    random_samples <- random_samples %>%
      mutate(upperBound = ((alpha - 1)*(log_x2 - log_x1)
                           - exp(log_x2 - log_x1))/2) %>%
      mutate(inRegion = log_x1 <= upperBound)
  
    accepted_samples <- random_samples %>%
      filter(inRegion == TRUE) %>%
      mutate(y = exp(log_x2 - log_x1))
    
    samples <- c(samples, accepted_samples$y)
    attempts <- attempts + n_remaining
    n_remaining <- n_remaining - nrow(accepted_samples)
  }
  return(list(samples=samples, attempts=attempts))
}

alphas <- seq(2,2000, 50)
rf_n <- partial(ratio_of_uniforms_f, n=1000)
attempts_df <- tibble(alpha=alphas, attempts = alphas %>% map_dbl(~rf_n(.x)$attempts))
attempts_df %>%
  ggplot(aes(x=alpha, y=attempts)) +
  geom_point() +
  geom_smooth()

accepted_samples <- ratio_of_uniforms_f(n, alpha)$samples %>%
  enframe() %>% 
  set_names(c('idx', 'y'))

accepted_samples %>%
  ggplot() +
  geom_histogram(aes(x=y, y=..density..), binwidth=0.2) +
  stat_function(aes(x=y), fun = dgamma, args = list(shape = alpha, scale = 1))

results <- list(mean=alpha,
                sample_mean=mean(accepted_samples$y),
                variance=alpha,
                sample_variance=var(accepted_samples$y))

print(results)
```






### 3. Arbitrary gamma function

```{r}
random_sample_from_f <- function(n, alpha){
  if (alpha < 1){
    result <- rejection_sampling_f(n, alpha)
  }
  else if (alpha == 1){
    result <- rexp(n, rate=1)$value
  }
  else{
    result <- ratio_of_uniforms_f(n, alpha)$samples
  }
  return(result)
}
```

