---
title: "Exercise 3 - Problem B"
author: "Jakob G. M. og Øyvind K."
output:
  ioslides_presentation: default
  beamer_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(broom)
library(tibble)

knitr::opts_chunk$set(echo = FALSE)
set.seed(0)

bilirubin <- here::here("reports", "report_3", "data", "bilirubin.txt") %>%
  read.table(header=T) %>%
  as_tibble()
```

## Overview of the data

Measurements of concentration of bilirubin in three individuals. Different number of measurements for each individual.

```{r}
bilirubin %>%
  group_by(pers) %>% 
  mutate(n = n()) %>% 
  distinct(pers, .keep_all=T) %>%
  select(pers, n) %>%
  knitr::kable(col.names = c("Person", "Number of measurements"))
```

## Regression model

$\log Y_{ij} = \beta_{i} + \epsilon_{ij}$,\ $i = 1,2,3$,\ $j = 1,\ldots,n_i$. 

$\epsilon_{ij} \sim \mathcal{N}(0, \sigma^2)$


## Boxplot of measurements across individuals

```{r}
bilirubin %>%
  mutate(log_meas = log(meas)) %>%
  ggplot(aes(x=pers, y=log_meas)) +
  geom_boxplot()
```



## Fitting a linear model

$$
H_0: \beta_1 = \beta_2 = \beta_3 \\
H_1: \beta_i \neq \beta_j\text{ for some }i \neq j
$$


```{r}
linear_model <- lm(log(meas)~pers, data=bilirubin)

Fval <- linear_model %>%
  glance() %>%
  pull(statistic)

linear_model %>%
  glance() %>%
  select(statistic, p.value) %>% 
  knitr::kable(col.names=c("F-statistic", "p-value"))
```

This leads to rejection of the null hypothesis.


## Permutation test

- Randomly assign measurements to individuals. 
- Fit linear model.
- Calculate the F-statistic.
- Repeat $N$ times.

$$
H_0:\text{ All measurements come from the same distribution} \\
H_1:\text{ The measurements are not from the same distribution}
$$
Reject $H_0$ if the F-statistic of the original data is larger than $95\%$ of F-statistics obtained from permuting the data. 

## Implementation of the permutation test

```{r, echo=TRUE}
permTest <- function(df) {
  perm_df <- tibble(
    meas = df %>%
      pull(meas) %>%
      sample(),
    pers = df %>%
      pull(pers)
  )
  lm(log(meas) ~ pers, data = perm_df) %>%
    glance() %>%
    pull(statistic)
}
```

## Conducting the permutation test

```{r, echo=TRUE}
permTest_df <- tibble(
  run = seq(1, 999),
  F_stat = 999 %>%
    rerun(permTest(bilirubin))
) %>%
  unnest()
```

```{r}
permTest_df %>%
  head(2) %>%
  bind_rows(permTest_df %>%
  tail(1)) %>%
  knitr::kable()
```


## Results

```{r}
perc_95 <- permTest_df %>%
  pull(F_stat) %>% 
  quantile(probs = c(0.95))

permTest_df %>%
  ggplot(aes(x=F_stat)) +
  geom_histogram(bins=50) +
  geom_vline(aes(xintercept=perc_95), color="red") +
  geom_vline(aes(xintercept=Fval), color="blue") +
  annotate("text", x = 2, y= 150, color="red", label="95th percentile") +
  annotate("text", x = 5.5, y= 150, color="blue", label="Original F-statistic")
  

p_value <- permTest_df %>%
  pull(F_stat) %>%
  map_lgl(~.x >= Fval) %>%
  mean()
```

## Results

$$
\text{Original F-statistic: }`r Fval` \\
\text{95th percentile: }  `r perc_95` \\
\text{P-value: } `r p_value`
$$

P-value is less than $0.05$, so we reject at significance level $\alpha = 0.05$. The original F-statistic is larger than the 95th percentile of F-statistics obtained from permuting the data. 

## Histogram and 95th percentile as $N$ grows

```{r, fig.height=2.2, fig.width=4, warning=FALSE}
N <- 2000
frames <- 50

permTest_df <- tibble(
  run = seq(1, N),
  F_stat = N %>%
    rerun(permTest(bilirubin))
) %>%
  unnest()

anim_df <- seq(0, frames) %>%
  map_dfr(~permTest_df %>%
            slice(0:(N/frames * (.x + 1))) %>%
            mutate(
              frame = .x + 1,
              perc_95 = F_stat %>%
                quantile(probs = c(0.95))))

binwidth = 0.2
anim_df %>%
  ggplot(aes(y = ..density..* binwidth, x=F_stat)) +
  geom_histogram(binwidth = binwidth) +
  geom_vline(aes(xintercept=perc_95), color="red") + 
  geom_vline(aes(xintercept=Fval), color="blue") +
  ylim(0, 0.2) +
  xlim(0, 8) +
  ylab("Density") + 
  xlab("F-statistic") + 
  annotate("text", x = 2, y= 0.2, color="red", label="95th percentile") +
  annotate("text", x = 5.5, y= 0.2, color="blue", label="Original F-statistic") +
  ggtitle("Number of permutations: {frame * N/nframes}") + 
  gganimate::transition_states(frame, transition_length = 1, state_length = 1)
```

